---
title: "Exercise #N"
author: "Your Name"
date: today
format:
  pdf:
    geometry: margin=1in
    fig-pos: "H"
    include-in-header:
      text: \usepackage{float}
---

<!--
HOW TO USE THIS FILE

Copy it, rename it, change the title and author, delete these comments, and
write. Everything below the setup chunk is placeholder text to write over.

Rendering it the first time will tell you whether your machine is set up:

  quarto render yourfile.qmd

If you get "No LaTeX installation found," you have two choices.

  1. Install one, once:            quarto install tinytex
  2. Skip LaTeX entirely: replace the whole format: block above with

       format:
         typst:
           papersize: us-letter
           fig-format: png
           margin:
             x: 1in
             y: 1in

     Typst ships inside Quarto, so there is nothing to install. It has no
     floating figures, so it does not need the fig-pos or float lines. The
     fig-format: png line is not optional -- Typst asks R for SVG figures by
     default, and R cannot make SVG on a Mac without XQuartz installed. You
     will know you have hit this if the render stops with "dev.control()
     called without an open graphics device."

FOR A PAPER RATHER THAN AN EXERCISE, add these to the pdf: block --

    fontsize: 12pt
    linestretch: 2          # double-spaced
    number-sections: true

-- and these at the top level, alongside title: and author: --

    abstract: |
      One paragraph.
    bibliography: refs.bib
    link-citations: true
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE,
                      fig.align = "center", fig.width = 6.5, fig.height = 4)

# print() wraps R output at this width; the default runs off the right margin
options(width = 70)

library(tidyverse)
```

# Setting up

Write prose in the body of the document, outside the chunks. Put a heading on
each section with `#`, and a subsection with `##`.

Replace the data below with your own. This is a placeholder so the template
renders on any machine without needing a file.

```{r data}
set.seed(20260918)

d <- tibble(x = rnorm(200),
            z = rbinom(200, 1, 0.5),
            y = 1.5 + 0.8 * x - 0.5 * z + rnorm(200))
```

# Results

Fit the model and show the table.

```{r model}
m <- lm(y ~ x + z, data = d)
summary(m)
```

Refer to a number from the model in your prose by putting R inside backticks
rather than typing it by hand, so it updates when the model does. The
coefficient on `x` is `r round(coef(m)["x"], 3)`.

```{r}
#| label: fig-fit
#| fig-cap: "Fitted values against x."

ggplot(d, aes(x = x, y = y)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", formula = y ~ x) +
  theme_minimal()
```

Give a figure a `label:` starting with `fig-` and a `fig-cap:`, then refer to
it as @fig-fit and the number takes care of itself.

# Notes

Three things that go wrong in PDF and not in HTML:

- **Long code lines get cut off at the right margin.** LaTeX will not wrap
  them. Keep source lines under about 80 characters and break long pipes
  across lines.
- **Figures drift** to the top of the next page unless pinned. The `fig-pos`
  and `float` lines in the header do that.
- **Package startup messages print into the document** unless suppressed. The
  `message = FALSE` in the setup chunk handles it.

To hide code but keep its output, put `#| echo: false` at the top of a chunk.
To run code and show nothing, use `#| include: false`. To show code without
running it, use `#| eval: false`.
